---
title: "Data Manipulation with dplyr"
author: "Joddy"
date: "2024-02-16"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Open Lab 1 - data manipulation with dplyr

Today we will focus on using dplyr to manipulate data. Rather than focus on data cleaning tasks, today we'll talk about how to select rows and columns from your data, how to derive new variables from your existing ones, use groups with summarize() and the slice functions to answer questions about your dataset.

This tutorial is based on the Data Transformation chapter in R for Data Science by Hadley Wickham: https://r4ds.hadley.nz/data-transform

I highly recommend that book as a guide for teaching yourself data science. It covers data visualization, transformation (what we are doing today), data cleaning, and programming.

```{r}
# Get the data we will be working with today
# install.packages("nycflights13")
```

Load our dataset, flights that departed from NYC
```{r}
library(nycflights13)
library(tidyverse)
```

```{r}
# Look at the data dictionary for flights dataset
?flights

# Save the flight data into the environment
flights <- flights
# Look at the tibble (data frame for tidyverse)
flights

# We only get the first few columns. What's the best way to see all of them?

# See all 19 columns using View()
View(flights)
```

## Pipes: Using dplyr to transform our data

The dplyr package can handle most of our data transformation tasks. The following are true for all dplyr functions:
  1. The first argument is always a data frame.
  2. The subsequent arguments typically describe which columns to operate on, using the variable names (without quotes).
  3. The output is always a new data frame.

The functions in dplyr are "verbs", like filter, group by, summarize, arrange. To solve complex problems, we need to combine multiple verbs. We will accomplish this with the pipe operator: |> or %>%.

'|>' is the built in R pipe. We will be using this one today, but %>% will work in place of it.

You will see '%>%' in older tidyverse code and tutorials, and it works just as well.

(It can do a few things that the base pipe can't, but for our purposes functions the same. For more details, see https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/)

The pipe works from left to right: the functions piped together will run starting at the left, and run in order until reaching the last item on the right side. This is the opposite direction from <-, so it's important to keep this in mind.

Here's an example of the pipe in action. We are using a series of verbs on the flights dataset to get the mean arrival delay for flights going from NYC to IAH (Houston airport), by each day. The rows have year, month, day, and the mean arrival delay for that day going from NYC to Houston.

```{r}
flights |>
  filter(dest == "IAH") |> 
  group_by(year, month, day) |> 
  summarize(
    arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```

- The dataset we are working with is "flights", so we start the line with that variable. The rest of the line is piped functions operating on that dataset.
- The first operation is "filter(dest == "IAH")", which gives us the rows for the Houston airport.
- Next, we group our data by year, month, day to get a mean for each day using "group_by(year, month, day)".
- Then, we use "summarize(arr_delay = mean(arr_delay, na.rm = TRUE))" to calculate the mean arrival delay for each day.

## Verbs that operate on rows

### 1. filter()
Filter changes which rows are present in a dataset based on value(s) in column(s). The argument is the condition on a column that must be true to keep the row. In the example, we had dest == "IAH", meaning that the destination was equal to "IAH" for Houston.

Notice that the "==" is required instead of "=" -- you'll get a warning and the filter won't work right if you use just one equals sign. We use == for conditions, and = for assigning a value to a variable.

#### Condition operators
  - You can use the familiar < and > for less than and greater than.
  - You can also do <= or >= for less than or equal to and greater than or equal to.
  - The operator != is for not equal to, which is useful for excluding rows with a certain value.
  - To combine two required conditions, use & (and).
  - To check for either of two conditions, use | (or).

```{r}
# Flights that departed on January 1
flights |> 
  filter(month == 1 & day == 1)

# Flights that departed in January or February
flights |> 
  filter(month == 1 | month == 2)
```

#### Saving filter results
```{r}
# filter() doesn't save the results to a data frame variable by default. The commands above print the results to the screen.
# To save to a data frame variable, you'll need to use <-.
jan1 <- flights |> 
  filter(month == 1 & day == 1)
```

### 2. arrange()
The arrange() function works like a sort, changing the order of the rows based on values in particular columns. It takes a data frame, and the column names you want to sort by. If you include more than one column, it will sort by the first one, then the second one where there are ties in the first variable, and so on. By default, it sorts in ascending order (we will see how to reverse the order in a minute).

```{r}
# For the following sort, we get earliest years first, then within a year the earliest months, and so on.
flights |> 
  arrange(year, month, day, dep_time)
```

```{r}
# To reverse the order, use the desc() function on a column. This will sort in descending order.
flights |> 
  arrange(desc(dep_delay))
```

### 3. Distinct
Find the unique rows in a dataset (remove duplicate rows). This step is very useful in data cleaning. The following removes any rows that are total duplicates (every column value for the rows are the same).
```{r}
flights |> 
  distinct()
```

If you supply arguments, the combination of the columns will be checked for duplicates, and only unique rows for those columns will be returned. For example to find all unique origin and destination pairs:
```{r}
flights |> 
  distinct(origin, dest)
```

This only gives you the origin and dest columns. To include the other columns too, use .keep_all = TRUE:
```{r}
flights |> 
  distinct(origin, dest, .keep_all = TRUE)
```

To get counts for occurences of column combinations, you can use count():
```{r}
flights |>
  count(origin, dest, sort = TRUE)
```

## Verbs that operate on columns

### mutate()
The mutate function creates new columns derived from existing columns. There are a large set of functions available to manipulate variables for your calculations. For now, we will stick with algebraic operators to illustrate how mutate works ( * / + - ).

For the flights data, we can use algebra to compute the gain, or how much time a delayed flight made up in the air.

This is found by subtracting the arrival delay from the departure delay, or:
  dep_delay - arr_delay

We can also use mutate() to find the speed of the aircraft, using: distance / air_time * 60

```{r}
flights <- flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )

View(flights)
```

Mutate will add the calculated columns (gain and speed) to the right of the existing columns. There are some options to change where the column goes.

  .before = [col # or col name] -  move the new column to the left of the column given.
  .after = [col # or col name] - move the new column to the right of the column given.

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )
```

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .after = day
  )
```

Sometimes we only want to keep certain columns. To only keep the columns used in the calculation:
```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

This will not overwrite the flights data frame, but will print the results to the screen. To overwrite it, you'll need to assign the result to a new variable:
```{r}
flights_mutated <- flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

Our new data frame only has the variables we created, and the ones used to perform the calculation.

Also notice that the pipe works left to right (from flights to mutate), but the assignment operator works from right to left (from flights to flights_mutated). This will get clearer with practice!

### select()

Select allows you to zoom in on just the variables you need to work with. There are a few ways to select columns:

#### By name
```{r}
flights |> 
  select(year, month, day)
```

#### Between columns (inclusive)
```{r}
flights |> 
  select(year:day)
# This will select all columns between year and day (including year and day).

# Excluding a range of columns (inclusive):
flights |> 
  select(!year:day)
```

This excludes all columns from year to day (including year and day).

See ?select for more details. There are helper functions providing more more ways to select columns!
```{r}
?select
```

#### relocate()

Relocate helps you move columns around. By default, it will move the column to the front (the far left):
```{r}
flights |> 
  relocate(time_hour, air_time)
```

You can specify where to put them using .before and .after like we did for mutate:
```{r}
flights |> 
  relocate(year:dep_time, .after = time_hour)
# Moves the columns from year to dep_time after the time_hour column

# Or, to move functions with a certain prefix:
flights |> 
  relocate(starts_with("arr"), .before = dep_time)
# All columns with the prefix "arr", the arrival columns, will go before the dep_time column.
```

## Piping with multiple verbs

Piping with multiple verbs is very powerful, and you can achieve a lot with one line of code. For example, imagine that you wanted to find the fast flights to Houston’s IAH airport: you need to combine filter(), mutate(), select(), and arrange():
```{r}
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```

As you read this, notice that there are 4 verbs that come in order: filter, mutate, select, and arrange. This tells us that we are:
  - filtering on some rows with a condition
  - calculating a speed variable
  - then selecting columns (including our newly calculated variable)
  - and then sorting by the speed variable.

This is a lot of work! But it can be written in one line of code.

What if we didn't have a pipe? How would we solve the problem to find the fastest flights?

We would have to nest the functions together like this:
```{r}
arrange(
  select(
    mutate(
      filter(
        flights, 
        dest == "IAH"
      ),
      speed = distance / air_time * 60
    ),
    year:day, dep_time, carrier, flight, speed
  ),
  desc(speed)
)
```

This is hard to read and understand!

Or, we could assign each result to a variable and move to the next step:
```{r}
flights1 <- filter(flights, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))
```

This is a little better, but still takes longer and is a bit harder to understand.

These other two methods are valid R code, but the pipe example is 1) short 2) easy to read 3) clear.

## Using groups

Groups let you group your data by columns. These groups don't change the underlying data, but help you to make summary calculations on your data based on a group. So you could look up the mean flight times by month, for example. The summarize() function lets us create variables to get summaries for the groups, like mean(), n(), sd(), and others.

First, use group_by() to create month groups:
```{r}
flights |> 
  group_by(month)
```

Then we can get a summary of the data by the groups we just created. To find the average delay by month, we can do this:
```{r}
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay)
  )
```

What went wrong? There were some NA values (missing values) in the dep_delay column, so the averaging did not work.

We need to add the argument "na.rm = TRUE" to our mean calculation to take missing values out of the calculation:
```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE)
  )
```

You can create multiple summaries in one summarize() call. Let's use n(), which will give a count summary
```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    count = n()
  )
```

## Slices

There are five handy functions that allow you extract specific rows within each group:
  df |> slice_head(n = 1) takes the first row from each group.
  df |> slice_tail(n = 1) takes the last row in each group.
  df |> slice_min(x, n = 1) takes the row with the smallest value of column x.
  df |> slice_max(x, n = 1) takes the row with the largest value of column x.
  df |> slice_sample(n = 1) takes one random row.

To take more than 1 row, you can change n = 1 to something else. Instead of n = , you can use prop = 0.1 to select 10% of the rows in each group.

The following finds the flights that are most delayed upon arrival at each destination:
```{r}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```

Notice that we have 105 destinations, but 108 rows. Why is this?

slice_min() and slice_max() keep tied values, so we get an extra row if there is a tie for a group.

To change this, you can set with_ties = FALSE:
```{r}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1, with_ties = FALSE) |>
  relocate(dest)
```

Now the number of rows in our summary matches the total destinations.

## Grouping by multiple variables

You can create groups using more than one variable, such as for each year/month/day.
```{r}
daily <- flights |>  
  group_by(year, month, day)
daily
```

When you summarize a tibble grouped by more than one variable, each summary peels off the last group. dplyr will display a message showing you which groups are being used.
```{r}
daily_flights <- daily |> 
  summarize(n = n())

# To hide the message, you can explicitly tell summarize which groups to use:
daily_flights <- daily |> 
  summarize(
    n = n(), 
    .groups = "drop_last"
  )
daily_flights
```

## Ungrouping

Sometimes you need to remove the groups, to change how you are summarizing. You can do this with ungroup():
  flights |> ungroup()

Now what happens when we summarize?
```{r}
flights |> ungroup() |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    flights = n()
  )
```

We get a single summary row back because all of the rows are together in an ungrouped data frame.

## Resources (crtl-click or cmd-click to open the pages)
This open lab was adapted from the book R for Data Science: https://r4ds.hadley.nz/data-transform

Visit https://go.ncsu.edu/getdatahelp for help with data science. You can email us, chat with a Data Science Consultant, or book a 30-minute consultation with a Data Science Consultant.

To find out more about upcoming library workshops, visit https://lib.ncsu.edu/workshops.

The Data & Visualization workshops offered by DVS are at https://www.lib.ncsu.edu/workshops/category/data-and-visualization

Eval: Please take a few minutes to fill out the workshop eval before you leave: https://go.ncsu.edu/dvs-eval
