---
title: "Exercises"
author: "Joddy"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Load our flights data
```{r}
library(nycflights13)
library(tidyverse)
# Look at the data dictionary for flights dataset
?flights

# Save the flight data into the environment
flights <- flights

flights <- flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )
```

## Piping

In a single pipeline for each condition, find all flights that meet the condition:

### 1. Had an arrival delay of two or more hours


### 2. Flew to Houston (IAH or HOU)


### 3. Were operated by United, American, or Delta


### 4. Departed in summer (July, August, and September)


### 5. Arrived more than two hours late, but didn’t leave late


### 6. Were delayed by at least an hour, but made up over 30 minutes in flight

What are some other interesting ways we could filter the flight data? How would you build the filter conditions for that?


Please share your ideas with the group!


## Selection

### 1. What happens if you specify the name of the same variable multiple times in a select() call?


### 2. Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?

```{r}
flights |> select(contains("TIME"))
```

# 3. Why doesn’t the following work, and what does the error mean?

```{r}
flights |> 
  select(tailnum) |> 
  arrange(arr_delay)
```

Output is:
  Error in `arrange()`:
  In argument: `..1 = arr_delay`.
  Caused by error:
  ! object 'arr_delay' not found
 
## Grouping

### 1. Which carrier has the worst average delays?

### 2. Challenge: can you disentangle the effects of bad airports vs. bad carriers?

Why/why not? (Hint: think about flights |> group_by(carrier, dest) |> summarize(n()))

Suppose we have the following tiny data frame:
```{r}  
  df <- tibble(
    x = 1:5,
    y = c("a", "b", "a", "a", "b"),
    z = c("K", "K", "L", "L", "K")
  )
```

### 1. Write down what you think the output will look like, then check if you were correct, and describe what group_by() does.
```{r}
df |>
  group_by(y)
```

### 2. Write down what you think the output will look like, then check if you were correct, and describe what arrange() does.
Also comment on how it’s different from the group_by() in part (a)?
```{r}  
  df |>
  arrange(y)
```

### 3. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.

```{r}
df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
```

### 4. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.
Then, comment on what the message says.

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```

# 5. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.

How is the output different from the one in part (d).

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
```

### 6. Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does.

How are the outputs of the two pipelines different?

```{r}  
  df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))

df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```