---
title: "Solutions"
author: "Joddy"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Solutions to exercises

## Load our flights data
```{r}
library(nycflights13)
library(tidyverse)
# Look at the data dictionary for flights dataset
?flights

# Save the flight data into the environment
flights <- flights

flights <- flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )
```

## Piping

In a single pipeline for each condition, find all flights that meet the condition:


### 1. Had an arrival delay of two or more hours
```{r}
flights |>
  filter(arr_delay >= 120)
```

### 2. Flew to Houston (IAH or HOU)
```{r}
flights |>
  filter(dest %in% c("IAH", "HOU"))
```

### 3. Were operated by United, American, or Delta
```{r}
flights |> filter(carrier == "UA" | carrier == "AA" | carrier == "DL")
flights |> filter(carrier %in% c("UA", "AA", "DL"))
```

### 4. Departed in summer (July, August, and September)
```{r}
flights |>
  filter(month %in% c(7,8,9))
```

### 5. Arrived more than two hours late, but didn’t leave late
```{r}
flights |>
  filter(arr_delay > 120 & dep_delay <= 0)
```

### 6. Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
flights |>
  filter(dep_delay >= 60 & gain > 30 )
```

What are some other interesting ways we could filter the flight data? How would you build the filter conditions for that?

Please share your ideas with the group!

## Selection

### 1. What happens if you specify the name of the same variable multiple times in a select() call?
```{r}
flights |>
  select(dep_delay, year, month, day, dep_delay)
```

The variable only shows up once.

### 2. Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?
```{r}
flights |> select(contains("TIME"))

# The select is not case sensitive by default.

# Add ignore.case=FALSE argument to contains.

flights |> select(contains("TIME", ignore.case=FALSE))
```

### 3. Why doesn’t the following work, and what does the error mean?
```{r}
flights |> 
  select(tailnum, arr_delay) |> 
  arrange(arr_delay) |>
  select(tailnum)
```

Output is:
  Error in `arrange()`:
  In argument: `..1 = arr_delay`.
  Caused by error:
  ! object 'arr_delay' not found
 
The arr_delay is not selected, only tailnum, so we can't sort by it because it isn't in the data frame.

## Grouping

### 1. Which carrier has the worst average delays?

### 2. Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights |> group_by(carrier, dest) |> summarize(n()))

```{r}
flights |> group_by(carrier,dest) |> 
  summarize(
    delay = mean(arr_delay, na.rm=TRUE),
    n(),
    .groups="keep"
  )
```

Suppose we have the following tiny data frame:
  
```{r}
  df <- tibble(
    x = 1:5,
    y = c("a", "b", "a", "a", "b"),
    z = c("K", "K", "L", "L", "K")
  )
```

### 1. Write down what you think the output will look like, then check if you were correct, and describe what group_by() does.
```{r}
df |>
  group_by(y) |> summarize(n = n())
```

Creates a group for a and b, the two values in the y variable. There are 3 a's and 2 b's, so we get n=3 for a and n=2 for b.

### 2. Write down what you think the output will look like, then check if you were correct, and describe what arrange() does. Also comment on how it’s different from the group_by() in part (a)?
```{r}
  df |>
  arrange(y)
```
  
The result will be sorted by the values of y, with a first and b second. All of the rows will be present.

### 3. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.

```{r}
df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
```

The mean(x) gives us the mean for the a group, and for the b group. The pipeline first sets up the groups of y (a and b), and then the summarize runs on the grouped tibble.

### 4. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. Then, comment on what the message says.
```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```

We are summarizing, so the z grouping will get dropped. We get a group for each level of y with a different level of z, so a, K and a,L, then for b both values have K for z, so they will stay grouped together.

### 5. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.
How is the output different from the one in part (d).
```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
```

This is the default, so we get the same output, but without the message saying the z group was dropped.

### 6. Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does.
How are the outputs of the two pipelines different?
```{r}  
  df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```
  
For this one, we get the 3 groups described in 4. We are using summarize to get the mean of x, so it will operate on the groups we created.

```{r}
df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```

Mutate will take the mean on the unique combinations of y and z. So there will be one mean for row 1 (a and K), one for rows 2 and 5 (b and K), and one for 3 and 4 (a and L). The mean for the "b and K" rows and the "a and L" rows are calculated differently:
  - a and L: (3+4+/2)=3.5
  - b an K: (2+5)/2=3.5


